cardinalR: Generating interesting high-dimensional data structures

### 2025-06-24 paper-cardinalR.Rmd
Updated \frac

Branch $1$ is generated from $X_1 \sim U(0, 1)$, and $X_2 = 0.1X_1 + X_1^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This produces a gently upward-curving parabola in the right half-plane. Branch $2$ is generated from $X_{1} \sim U(-1, 0)$, and $X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This creates a steeper, leftward-facing curve in the left half-plane. Branches $3$ to $k$ are added iteratively. Each new branch $i$ begins at a randomly chosen point within a restricted horizontal band: $X_{1} \in [-0.15, 0.15]$ (to ensure connectivity with earlier branches), spans a unit-length interval on $X_{1}$:$X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]$, and has the structure:

$$
X_2 = 0.1 \cdot X_1 - s_i \cdot (X_1^2 - x_{\text{start}}) + y_{\text{start}}, \quad s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1.5\}.
$$

Here, $s_i$ is a sampled scale factor (slope-like term) controlling curvature, and $(x_{\text{start}}, y_{\text{start}})$ is the starting point from the existing structure.

This construction yields $k$ spatially connected, nonlinear branches with varying curvature.

For $p > 2$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

@@

For branch $i$, $n_i$ points are generated, where $\sum_{i=1}^k n_i = n$. The structure follows: $X_{i1} \sim U(0, 2)$, and $X_{i2} = -s_i \cdot x_{i1} + \epsilon, \quad \epsilon \sim U(0, 0.5)$, where $s_i$ is a scale factor controlling the slope. When $k \leq \binom{p}{2}$, $s_i = 1$. When sampling with replacement, $s_i$ is drawn from the set ${1, 1.5, 2, \dots, 8}$ in increments of $0.5$. For $p > 2$, the remaining dimensions contain independent Gaussian noise: $X_j \sim N(0, 0.1^2)$, for $j \notin \{i1, i2\}$.

@@

Each branch contains $n_i$ points such that $\sum_{i=1}^k n_i = n$, where the vector $(n_1, \dots, n_k)$ is randomly drawn using the helper function `gen_nsum()` to partition $n$. Within its assigned subspace $(X_{i1}, X_{i2})$, branch $i$ is defined by: $X_{i1} \sim U(0, 2)$, and $X_{i2} = -s_i \cdot X_{i1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)$. This forms a smooth downward-opening parabola in the plane defined by $(X_{i1}, X_{i2})$, with the degree of curvature controlled by $s_i$. For $p > 2$, all remaining dimensions are independent Gaussian noise: $X_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}$.

@@

The Mobius strip is generated using `geozoo::mobius(n, p = 3)$points`, which samples $n$ points from a parametrization of the surface defined by angle $\theta$ and strip width $w$. The cartesian coordinates are $X_1 = \left(1 + \frac{w}{2} \cos(\theta/2)\right) \cos(\theta)$, $X_2 = \left(1 + \frac{w}{2} \cos(\theta/2)\right) \sin(\theta)$, and $X_3 = \frac{w}{2} \sin(\theta/2)$. This maps a $2\text{-}D$ band with a half-twist into $3\text{-}D$ space, forming a non-orientable one-sided surface. For $p > 3$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $3\text{-}D$ Mobius into $p\text{-}D$, where $j = 4, \dots, p$.

@@

Let $X_1, X_2, \dots, X_p$ denote the coordinates of the generated points. The final dimension $X_p$ encodes the height of each point and is drawn from an exponential distribution capped at $h$: $X_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right)$. Let $r_x(z)$ and $r_y(z)$ denote the half-widths of the rectangular cross-section at height $z$: $r_x(z) = r_t + (l_x - r_t) \cdot \frac{z}{h}$, $r_y(z) = r_t + (l_y - r_t) \cdot \frac{z}{h}$. The first three coordinates are then defined as: $X_1 \sim U(-r_x(z),\ r_x(z))$, $X_2 \sim U(-r_y(z),\ r_y(z))$, and $X_3 \sim U(-r_x(z),\ r_x(z))$. For $p > 3$, the remaining $p - 4$ dimensions (i.e., $X_4$ to $X_{p-1}$) are tapered toward zero with decreasing height and generated as: $X_j \sim U(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right), \quad \text{for } j = 4, \dots, p-1$. This tapering ensures that the structure narrows in all directions as it approaches the tip.

@@

Let $X_1, X_2, \dots, X_p$ denote the coordinates of the generated points. The final coordinate $X_p$ encodes the height $z$ of each point and is drawn from an exponential distribution capped at $h$: $X_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right)$. Let $r(z)$ denote the scaling factor (distance from the origin to triangle vertices) at height $z$: $r(z) = r_t + (l - r_t) \cdot \frac{z}{h}$.

A point in the triangle at height $z$ is generated using barycentric coordinates $(u, v)$ to ensure uniform sampling within the triangular cross-section: $u, v \sim U(0, 1), \quad \text{if } u + v > 1: u \leftarrow 1 - u,\ v \leftarrow 1 - v$. The first three coordinates (triangle plane) are then: $X_1 = r(z)(1 - u - v)$, $X_2 = r(z) \cdot u$, and $X_3 = r(z) \cdot v$. For dimensions $j = 4, \dots, p-1$, values are tapered linearly toward zero near the tip: $X_j \sim U(-0.1, 0.1) \cdot \left(1 - \frac{z}{h}\right)$.

@@

The radius at height $z$ scales linearly from zero (tip) to the base radius $r_b$: $r(z) = r_b \cdot \left(1 - \frac{z}{h}\right)$.

Each point is placed within a regular hexagon in the plane $(X_1, X_2)$, using a randomly chosen hexagon sector angle $\theta \in \{0, \frac{\pi}{3}, \frac{2\pi}{3}, \pi, \frac{4\pi}{3}, \frac{5\pi}{3}\}$ and a uniformly random radial scaling factor: $\theta \sim \text{Uniform sample from 6 hexagon angles}$,
$r_{\text{point}} \sim \sqrt{U(0, 1)}$. Then, the first two coordinates are: $X_1 = r(z) \cdot r_{\text{point}} \cdot \cos(\theta)$, and $X_2 = r(z) \cdot r_{\text{point}} \cdot \sin(\theta)$. For dimensions $j = 3, \dots, p-1$, values are tapered toward zero near the tip: $X_j \sim U(-0.1, 0.1) \cdot \left(1 - \frac{z}{h} \right)$.

@@

Let $C_1, C_2, \dots, C_{p+1}$ denote the corner vertices of a $p\text{-}D$ simplex. At each iteration $i = 1, \dots, n$, a new point is computed by taking the midpoint between the previous point $T_{i-1}$ and a randomly selected vertex $C_k$: $T_i = \frac{1}{2}(T_{i-1} + C_k), \quad C_k \in \{C_1, \dots, C_{p+1}\}$. This recursive midpoint rule generates self-similar patterns with systematic voids (holes) between clusters of points. The points remain bounded inside the convex hull of the simplex. The final output is a $n \times p$ matrix where each row represents a point: $X = \{T_1, T_2, \dots, T_n\}, \quad X \in \mathbb{R}^{n \times p}$.

@@

The $3\text{-}D$ geometry is constructed by introducing a latent parameter, $\theta \sim U\left(-\frac{3\pi}{2}, \frac{3\pi}{2}\right)$. This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: $X_1 = \sin(\theta)$, $X_2 \sim U(0, 2)$, $X_3 = \text{sign}(\theta) \cdot (\cos(\theta) - 1)$. This configuration creates a horizontally curled shape in $(X_1, X_3)$, with additional band thickness in the $X_2$ direction. For $p > 3$, additional noise dimensions are appended introducing structured, wavy noise.

@@

A latent angle variable $\theta$ is uniformly sampled from the interval $[0, 2\pi]$. Coordinates in the first two dimensions represent a perfect circle on the plane: $X_1 = \cos(\theta), \quad X_2 = \sin(\theta)$. For dimensions $X_3$ through $X_p$, sinusoidal transformations of the angle $\theta$ are introduced. The first component is a scaling factor that decreases with the dimension index, defined as $\text{scale}_j = \sqrt{(0.5)^{j-2}}$ for $j = 3, \dots, p$. The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula $\phi_j = (j - 2) \cdot \frac{\pi}{2p}$. Each additional dimension is computed as: $X_j = \text{scale}_{j} \cdot \sin(\theta + \phi_j), \quad j = 3, \dots, p$.

@@

A latent angle variable $\theta$ is uniformly sampled from the interval $[0, 2\pi]$. The first three dimensions define a non-circular closed curve, referred to as a "curvy cycle." In this configuration, $X_1 = \cos(\theta)$ represents horizontal oscillation, while $X_2 = \sqrt{3}/3 + \sin(\theta)$ introduces a vertical offset to avoid centering the curve at the origin. Additionally, $X_3 = \frac{1}{3} \cdot \cos(3\theta)$ introduces a third harmonic perturbation that intricately folds the curve three times along its path, creating a unique and complex shape that oscillates in both dimensions while incorporating the effects of the harmonic perturbation.

Together, these define a periodic, non-trivial, closed curve in $3\text{-}D$ with internal folds that produce a more complex geometry than a standard circle or ellipse. For dimensions $X_4$ through $X_p$, additional structured variability is introduced through decreasing amplitude scaling and phase-shifted sine waves. The scaling factor is defined as $\text{scale}_j = \sqrt{(0.5)^{j-3}}$ for $j$ ranging from $4$ to $p$, which means that the amplitude decreases as the dimension increases. Each dimension $X_j$ is then calculated using the formula $X_j = \text{scale}_j \cdot \sin(\theta + \phi_j)$, where the phase shift $\phi_j$ is given by $\phi_j = (j - 2) \cdot \frac{\pi}{2p}$.

@@

Each observation lies on the surface of a sphere in $3\text{-}D$, constructed by generating $u \sim U(-1, 1)$ which represents the cosine of the polar angle $\phi$ and $\theta \sim U(0, 2\pi)$ which represents the azimuthal angle. The corresponding Cartesian coordinates are calculated as; $X_1 = r \cdot \sqrt{1 - u^2} \cdot \cos(\theta)$, $X_2 = r \cdot \sqrt{1 - u^2} \cdot \sin(\theta)$, $X_3 = r \cdot u$, which gives points uniformly distributed on the surface of a $3\text{-}D$ sphere (not within). For $p > 3$, additional dimensions $X_4$ to $X_p$ are generated as low-variance Gaussian noise: $X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p$.

@@

Each point on the sphere is computed using the spherical-to-Cartesian transformation: $X_1 = \sin(\phi) \cdot \cos(\theta)$, $X_2 = \sin(\phi) \cdot \sin(\theta)$, $X_3 = \cos(\phi)$. For $p > 3$, the manifold is embedded in a higher-dimensional space by adding Gaussian noise dimensions: $X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p$.

@@

Each observation is situated on a restricted $4\text{-}D$ spherical surface, defined by spherical coordinates. The azimuthal angle $\theta_1 \sim U(0, \pi)$ in the $(x_1, x_2)$ plane, while the elevation angle $\theta_2 \sim U(0, \pi)$ in the $(x_2, x_3)$ plane. Additionally, $\theta_3 \sim U(0, \pi/2)$ in the $(x_3, x_4)$ plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into $4\text{-}D$ Cartesian space: $X_1 = \sin(\theta_1) \cdot \cos(\theta_2)$, $X_2 = \sin(\theta_1) \cdot \sin(\theta_2)$, $X_3 = \cos(\theta_1) \cdot \cos(\theta_3)$, $X_4 = \cos(\theta_1) \cdot \sin(\theta_3)$. This produces points on one side of a $4\text{-}D$ unit sphere, effectively generating a $4\text{-}D$ hemisphere. For $p > 4$, additional dimensions $X_5$ to $X_p$ are added as low-variance Gaussian noise: $X_j \sim N(0, 0.05^2),\text{ for }j = 5, \dots, p$.

@@

The first two dimensions define a crescent-shaped curve using a semi-circular arc. Let $\theta \in [\pi/6, 2\pi]$ be a sequence of $n$ evenly spaced angles. The corresponding $2\text{-}D$ coordinates are defined by: $X_1 = \cos(\theta)$, $X_2 = \sin(\theta)$. For $p > 2$, additional dimensions $X_3$ to $X_p$ are generated using `gen_wavydims1()`: $X_j = s_j \cdot \theta + \varepsilon_j,\text{ where }\varepsilon_j \sim N(0, 0.5^2),\text{ and }s_j\text{ is a random scale}$. These noise dimensions are nonlinearly correlated with $\theta$, producing structured, wave-like variation that aligns with the progression along the crescent curve.

@@

Cartesian coordinates from spherical conversion: $X_1 = \sin(\phi) \cdot \cos(\theta)$, $X_2 = \sin(\phi) \cdot \sin(\theta)$, $X_3 = \cos(\phi) + \varepsilon$, where $\varepsilon \sim U(-0.5, 0.5)$ introduces vertical jitter, and $X_4 = \theta / \max(\theta)$: a normalized progression along the spiral path. This generates a spherical spiral curve embedded in $4\text{-}D$ space, combining both circular and vertical movement, with gentle curvature and non-linear progression.

If $p > 4$, the function appends structured, non-linear noise via `gen_wavydims2()`. These dimensions are functions of the first coordinate $X_1$, introducing dependencies that preserve some geometric coherence. Each added dimension follows the form: $X_j = s_j \cdot (-1)^{\lfloor j/2 \rfloor} \cdot X_1^{a_j} + \eta_j$, where $a_j \in \{2, 3, 4, 5\}$ is a randomly chosen polynomial power, $s_j \sim U(0.5, 2)$ is a scale factor, $\eta_j \sim U(-\sigma_j, 2\sigma_j)$, with $\sigma_j \sim U(0, 0.05)$, adds mild randomness.

@@

The first four coordinates follow a $3\text{-}D$ helix with an additional oscillating component. Let $\theta \in [0, \frac{5\pi}{4}]$ be a sequence of angles controlling rotation around a circle. Cartesian coordinates; $X_1 = \cos(\theta)$: circular trajectory along the x-axis, $X_2 = \sin(\theta)$: circular trajectory along the y-axis, $X_3 = 0.05 \cdot \theta + \varepsilon_3$, with $\varepsilon_3 \sim U(-0.5, 0.5)$: linear progression (height) with vertical jitter, simulating a helix, and $X_4 = 0.1 \cdot \sin(\theta)$: oscillates with $\theta$, representing a periodic "wobble" along the fourth dimension. This results in a helical spiral winding upward along the third axis, with gentle sinusoidal fluctuation in the fourth dimension. When $p > 4$, the remaining $p - 4$ dimensions are populated using `gen_noisedims()`. These are independent Gaussian noise dimensions: $X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p$.

@@

The shape is defined by parameter $\theta \in [0, 2\pi \cdot \text{spins}]$, controlling the angular progression of the spiral. The Archimedean spiral in the horizontal plane is represented by; $X_1 = \theta \cdot \cos(\theta)$ for radial expansion in x, and $X_2 = \theta \cdot \sin(\theta)$ for radial expansion in y. The growth pattern resembles a cone, with the height increasing according to $X_3 = 2 \cdot \theta / \max(\theta) + \varepsilon_3$, with $\varepsilon_3 \sim U(-0.1, 0.6).$ Spiral modulation in the fourth dimension is represented by $X_4 = \theta \cdot \sin(2\theta) + \varepsilon_4$, with $\varepsilon_4 \sim U(-0.1, 0.6)$ which simulates a twisting helical component in a non-radial dimension. This results in a $3\text{-}D$ spiral surface expanding upward and outward like a cone, with an oscillatory fourth dimension capturing spiral irregularities. For $p > 4$, the dataset includes isotropic Gaussian noise in the remaining dimensions via `gen_noisedims()`: $X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p$.

@@

The $X_{1} \sim U(0.1, 2)$: base variable (avoids zero to prevent division errors), $X_{3} \sim U(0.1, 0.8)$: independent auxiliary variable, $X_{2} = \frac{hc}{X_{1}} + \text{nonfac} \cdot \sin(X_{1})$: non-linear combination of hyperbolic and sinusoidal transformations, creating sharp curvature and oscillation, and $X_{4} = \cos(\pi \cdot X_{1}) + \varepsilon$, with $\varepsilon \sim U(-0.1, 0.1)$: additional nonlinear variation based on cosine, simulating more subtle periodic structure. These transformations together result in a non-linear surface warped in multiple ways: sharp vertical shifts due to inverse terms, smooth waves from sine and cosine, and additional jitter. If $p > 4$, the remaining dimensions are populated with Gaussian noise using `gen_noisedims()`: $X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p$.

>>>>

Branch $1$ is generated from $X_1 \sim U(0, 1)$, and $X_2 = 0.1X_1 + X_1^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This produces a gently upward-curving parabola in the right half-plane. Branch $2$ is generated from $X_{1} \sim U(-1, 0)$, and $X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This creates a steeper, leftward-facing curve in the left half-plane. Branches $3$ to $k$ are added iteratively. Each new branch $i$ begins at a randomly chosen point within a restricted horizontal band: $X_{1} \in [-0.15, 0.15]$ (to ensure connectivity with earlier branches), spans a unit-length interval on $X_{1}$:$X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]$, and has the structure: $X_2 = 0.1X_1 - s_i(X_1^2 - x_{\text{start}}) + y_{\text{start}}, s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1.5\}$. Here, $s_i$ is a sampled scale factor (slope-like term) controlling curvature, and $(x_{\text{start}}, y_{\text{start}})$ is the starting point from the existing structure. This construction yields $k$ spatially connected, nonlinear branches with varying curvature. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

@@

For branch $i$, $n_i$ points are generated, where $\sum_{i=1}^k n_i = n$. The structure follows: $X_{i1} \sim U(0, 2)$, and $X_{i2} = -s_iX_{i1} + \epsilon, \quad \epsilon \sim U(0, 0.5)$, where $s_i$ is a scale factor controlling the slope. When $k \leq \binom{p}{2}$, $s_i = 1$. When sampling with replacement, $s_i$ is drawn from the set ${1, 1.5, 2, \dots, 8}$ in increments of $0.5$. For $p > 2$, the remaining dimensions contain independent Gaussian noise: $X_j \sim N(0, 0.1^2)$, for $j \notin \{i1, i2\}$.

@@

Each branch contains $n_i$ points such that $\sum_{i=1}^k n_i = n$, where the vector $(n_1, \dots, n_k)$ is randomly drawn using the helper function `gen_nsum()` to partition $n$. Within its assigned subspace $(X_{i1}, X_{i2})$, branch $i$ is defined by: $X_{i1} \sim U(0, 2)$, and $X_{i2} = -s_iX_{i1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)$. This forms a smooth downward-opening parabola in the plane defined by $(X_{i1}, X_{i2})$, with the degree of curvature controlled by $s_i$. For $p > 2$, all remaining dimensions are independent Gaussian noise: $X_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i1, i2\}$.

@@

The Mobius strip is generated using `geozoo::mobius(n, p = 3)$points`, which samples $n$ points from a parametrization of the surface defined by angle $\theta$ and strip width $w$. The cartesian coordinates are $X_1 = \left(1 + w/2 \cos(\theta/2)\right) \cos(\theta)$, $X_2 = \left(1 + w/2 \cos(\theta/2)\right) \sin(\theta)$, and $X_3 = w/2 \sin(\theta/2)$. This maps a $2\text{-}D$ band with a half-twist into $3\text{-}D$ space, forming a non-orientable one-sided surface. For $p > 3$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $3\text{-}D$ Mobius into $p\text{-}D$, where $j = 4, \dots, p$.

@@

Let $X_1, X_2, \dots, X_p$ denote the coordinates of the generated points. The final dimension $X_p$ encodes the height of each point and is drawn from an exponential distribution capped at $h$: $X_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right)$. Let $r_x(z)$ and $r_y(z)$ denote the half-widths of the rectangular cross-section at height $z$: $r_x(z) = r_t + (l_x - r_t)z/h$, $r_y(z) = r_t + (l_y - r_t)z/h$. The first three coordinates are then defined as: $X_1 \sim U(-r_x(z),\ r_x(z))$, $X_2 \sim U(-r_y(z),\ r_y(z))$, and $X_3 \sim U(-r_x(z),\ r_x(z))$. For $p > 3$, the remaining $p - 4$ dimensions (i.e., $X_4$ to $X_{p-1}$) are tapered toward zero with decreasing height and generated as: $X_j \sim U(-0.1, 0.1)\left(1 - z/h\right), \quad \text{for } j = 4, \dots, p-1$. This tapering ensures that the structure narrows in all directions as it approaches the tip.

@@

Let $X_1, X_2, \dots, X_p$ denote the coordinates of the generated points. The final coordinate $X_p$ encodes the height $z$ of each point and is drawn from an exponential distribution capped at $h$: $X_p = z \sim \min\left(\text{Exp}(\lambda = 2/h),\ h\right)$. Let $r(z)$ denote the scaling factor (distance from the origin to triangle vertices) at height $z$: $r(z) = r_t + (l - r_t)z/h$.

A point in the triangle at height $z$ is generated using barycentric coordinates $(u, v)$ to ensure uniform sampling within the triangular cross-section: $u, v \sim U(0, 1), \quad \text{if } u + v > 1: u \leftarrow 1 - u,\ v \leftarrow 1 - v$. The first three coordinates (triangle plane) are then: $X_1 = r(z)(1 - u - v)$, $X_2 = r(z)u$, and $X_3 = r(z)v$. For dimensions $j = 4, \dots, p-1$, values are tapered linearly toward zero near the tip: $X_j \sim U(-0.1, 0.1)\left(1 - z/h\right)$.

@@

The radius at height $z$ scales linearly from zero (tip) to the base radius $r_b$: $r(z) = r_b\left(1 - z/h\right)$.

Each point is placed within a regular hexagon in the plane $(X_1, X_2)$, using a randomly chosen hexagon sector angle $\theta \in \{0, \pi/3, 2\pi/3, \pi, 4\pi/3, 5\pi/3\}$ and a uniformly random radial scaling factor: $\theta \sim \text{Uniform sample from 6 hexagon angles}$,
$r_{\text{point}} \sim \sqrt{U(0, 1)}$. Then, the first two coordinates are: $X_1 = r(z)r_{\text{point}}\cos(\theta)$, and $X_2 = r(z)r_{\text{point}}\sin(\theta)$. For dimensions $j = 3, \dots, p-1$, values are tapered toward zero near the tip: $X_j \sim U(-0.1, 0.1)\left(1 - z/h \right)$.

@@

Let $C_1, C_2, \dots, C_{p+1}$ denote the corner vertices of a $p\text{-}D$ simplex. At each iteration $i = 1, \dots, n$, a new point is computed by taking the midpoint between the previous point $T_{i-1}$ and a randomly selected vertex $C_k$: $T_i = 1/2(T_{i-1} + C_k), \quad C_k \in \{C_1, \dots, C_{p+1}\}$. This recursive midpoint rule generates self-similar patterns with systematic voids (holes) between clusters of points. The points remain bounded inside the convex hull of the simplex. The final output is a $n \times p$ matrix where each row represents a point: $X = \{T_1, T_2, \dots, T_n\}, \quad X \in \mathbb{R}^{n \times p}$.

@@

The $3\text{-}D$ geometry is constructed by introducing a latent parameter, $\theta \sim U\left(-3\pi/2, 3\pi/2\right)$. This parameter controls the curvature of the manifold. The first three dimensions form the S-curve structure: $X_1 = \sin(\theta)$, $X_2 \sim U(0, 2)$, $X_3 = \text{sign}(\theta)(\cos(\theta) - 1)$. This configuration creates a horizontally curled shape in $(X_1, X_3)$, with additional band thickness in the $X_2$ direction. For $p > 3$, additional noise dimensions are appended introducing structured, wavy noise.

@@

A latent angle variable $\theta$ is uniformly sampled from the interval $[0, 2\pi]$. Coordinates in the first two dimensions represent a perfect circle on the plane: $X_1 = \cos(\theta), \quad X_2 = \sin(\theta)$. For dimensions $X_3$ through $X_p$, sinusoidal transformations of the angle $\theta$ are introduced. The first component is a scaling factor that decreases with the dimension index, defined as $\text{scale}_j = \sqrt{(0.5)^{j-2}}$ for $j = 3, \dots, p$. The second component is a phase shift that is proportional to the dimension index, specifically designed to decorrelate the curves, given by the formula $\phi_j = (j - 2)\pi/2p$. Each additional dimension is computed as: $X_j = \text{scale}_{j}\sin(\theta + \phi_j), \quad j = 3, \dots, p$.

@@

A latent angle variable $\theta$ is uniformly sampled from the interval $[0, 2\pi]$. The first three dimensions define a non-circular closed curve, referred to as a "curvy cycle." In this configuration, $X_1 = \cos(\theta)$ represents horizontal oscillation, while $X_2 = \sqrt{3}/3 + \sin(\theta)$ introduces a vertical offset to avoid centering the curve at the origin. Additionally, $X_3 = 1/3\cos(3\theta)$ introduces a third harmonic perturbation that intricately folds the curve three times along its path, creating a unique and complex shape that oscillates in both dimensions while incorporating the effects of the harmonic perturbation.

Together, these define a periodic, non-trivial, closed curve in $3\text{-}D$ with internal folds that produce a more complex geometry than a standard circle or ellipse. For dimensions $X_4$ through $X_p$, additional structured variability is introduced through decreasing amplitude scaling and phase-shifted sine waves. The scaling factor is defined as $\text{scale}_j = \sqrt{(0.5)^{j-3}}$ for $j$ ranging from $4$ to $p$, which means that the amplitude decreases as the dimension increases. Each dimension $X_j$ is then calculated using the formula $X_j = \text{scale}_j\sin(\theta + \phi_j)$, where the phase shift $\phi_j$ is given by $\phi_j = (j - 2)\pi/2p$.

@@

Each observation lies on the surface of a sphere in $3\text{-}D$, constructed by generating $u \sim U(-1, 1)$ which represents the cosine of the polar angle $\phi$ and $\theta \sim U(0, 2\pi)$ which represents the azimuthal angle. The corresponding Cartesian coordinates are calculated as; $X_1 = r\sqrt{1 - u^2}\cos(\theta)$, $X_2 = r\sqrt{1 - u^2}\sin(\theta)$, $X_3 = ru$, which gives points uniformly distributed on the surface of a $3\text{-}D$ sphere (not within). For $p > 3$, additional dimensions $X_4$ to $X_p$ are generated as low-variance Gaussian noise: $X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p$.

@@

Each point on the sphere is computed using the spherical-to-Cartesian transformation: $X_1 = \sin(\phi)\cos(\theta)$, $X_2 = \sin(\phi)\sin(\theta)$, $X_3 = \cos(\phi)$. For $p > 3$, the manifold is embedded in a higher-dimensional space by adding Gaussian noise dimensions: $X_j \sim N(0, 0.05^2),\text{ for }j = 4, \dots, p$.

@@

Each observation is situated on a restricted $4\text{-}D$ spherical surface, defined by spherical coordinates. The azimuthal angle $\theta_1 \sim U(0, \pi)$ in the $(x_1, x_2)$ plane, while the elevation angle $\theta_2 \sim U(0, \pi)$ in the $(x_2, x_3)$ plane. Additionally, $\theta_3 \sim U(0, \pi/2)$ in the $(x_3, x_4)$ plane, ensuring that the points remain restricted to a hemisphere. The coordinates are transformed into $4\text{-}D$ Cartesian space: $X_1 = \sin(\theta_1)\cos(\theta_2)$, $X_2 = \sin(\theta_1)\sin(\theta_2)$, $X_3 = \cos(\theta_1)\cos(\theta_3)$, $X_4 = \cos(\theta_1)\sin(\theta_3)$. This produces points on one side of a $4\text{-}D$ unit sphere, effectively generating a $4\text{-}D$ hemisphere. For $p > 4$, additional dimensions $X_5$ to $X_p$ are added as low-variance Gaussian noise: $X_j \sim N(0, 0.05^2),\text{ for }j = 5, \dots, p$.

@@

The first two dimensions define a crescent-shaped curve using a semi-circular arc. Let $\theta \in [\pi/6, 2\pi]$ be a sequence of $n$ evenly spaced angles. The corresponding $2\text{-}D$ coordinates are defined by: $X_1 = \cos(\theta)$, $X_2 = \sin(\theta)$. For $p > 2$, additional dimensions $X_3$ to $X_p$ are generated using `gen_wavydims1()`: $X_j = s_j\theta + \varepsilon_j,\text{ where }\varepsilon_j \sim N(0, 0.5^2),\text{ and }s_j\text{ is a random scale}$. These noise dimensions are nonlinearly correlated with $\theta$, producing structured, wave-like variation that aligns with the progression along the crescent curve.

@@

Cartesian coordinates from spherical conversion: $X_1 = \sin(\phi)\cos(\theta)$, $X_2 = \sin(\phi)\sin(\theta)$, $X_3 = \cos(\phi) + \varepsilon$, where $\varepsilon \sim U(-0.5, 0.5)$ introduces vertical jitter, and $X_4 = \theta / \max(\theta)$: a normalized progression along the spiral path. This generates a spherical spiral curve embedded in $4\text{-}D$ space, combining both circular and vertical movement, with gentle curvature and non-linear progression.

If $p > 4$, the function appends structured, non-linear noise via `gen_wavydims2()`. These dimensions are functions of the first coordinate $X_1$, introducing dependencies that preserve some geometric coherence. Each added dimension follows the form: $X_j = s_j(-1)^{\lfloor j/2 \rfloor}X_1^{a_j} + \eta_j$, where $a_j \in \{2, 3, 4, 5\}$ is a randomly chosen polynomial power, $s_j \sim U(0.5, 2)$ is a scale factor, $\eta_j \sim U(-\sigma_j, 2\sigma_j)$, with $\sigma_j \sim U(0, 0.05)$, adds mild randomness.

@@

The first four coordinates follow a $3\text{-}D$ helix with an additional oscillating component. Let $\theta \in [0, 5\pi/4]$ be a sequence of angles controlling rotation around a circle. Cartesian coordinates; $X_1 = \cos(\theta)$: circular trajectory along the x-axis, $X_2 = \sin(\theta)$: circular trajectory along the y-axis, $X_3 = 0.05\theta + \varepsilon_3$, with $\varepsilon_3 \sim U(-0.5, 0.5)$: linear progression (height) with vertical jitter, simulating a helix, and $X_4 = 0.1\sin(\theta)$: oscillates with $\theta$, representing a periodic "wobble" along the fourth dimension. This results in a helical spiral winding upward along the third axis, with gentle sinusoidal fluctuation in the fourth dimension. When $p > 4$, the remaining $p - 4$ dimensions are populated using `gen_noisedims()`. These are independent Gaussian noise dimensions: $X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p$.

@@

The shape is defined by parameter $\theta \in [0, 2\pi\text{spins}]$, controlling the angular progression of the spiral. The Archimedean spiral in the horizontal plane is represented by; $X_1 = \theta\cos(\theta)$ for radial expansion in x, and $X_2 = \theta\sin(\theta)$ for radial expansion in y. The growth pattern resembles a cone, with the height increasing according to $X_3 = 2\theta / \max(\theta) + \varepsilon_3$, with $\varepsilon_3 \sim U(-0.1, 0.6).$ Spiral modulation in the fourth dimension is represented by $X_4 = \theta\sin(2\theta) + \varepsilon_4$, with $\varepsilon_4 \sim U(-0.1, 0.6)$ which simulates a twisting helical component in a non-radial dimension. This results in a $3\text{-}D$ spiral surface expanding upward and outward like a cone, with an oscillatory fourth dimension capturing spiral irregularities. For $p > 4$, the dataset includes isotropic Gaussian noise in the remaining dimensions via `gen_noisedims()`: $X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p$.

@@

The $X_{1} \sim U(0.1, 2)$: base variable (avoids zero to prevent division errors), $X_{3} \sim U(0.1, 0.8)$: independent auxiliary variable, $X_{2} = hc/X_{1} + \text{nonfac}\sin(X_{1})$: non-linear combination of hyperbolic and sinusoidal transformations, creating sharp curvature and oscillation, and $X_{4} = \cos(\pi X_{1}) + \varepsilon$, with $\varepsilon \sim U(-0.1, 0.1)$: additional nonlinear variation based on cosine, simulating more subtle periodic structure. These transformations together result in a non-linear surface warped in multiple ways: sharp vertical shifts due to inverse terms, smooth waves from sine and cosine, and additional jitter. If $p > 4$, the remaining dimensions are populated with Gaussian noise using `gen_noisedims()`: $X_j \sim N(0, 0.05^2), \quad j = 5, \dots, p$.
